---
title: "Week_06_Hip_Replacement_Exercise"
author: "Catriona Maclean"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aim 

To analyse the data on hip replacements and answer the following questions.

1. Plot 'EQ-5D Index' scores (a combination of five key criteria concerning patients' self-reported general health) pre and post operation for each gender

2. Calculate how many patients in this dataset have been told by a doctor that they have problems caused by a stroke

3. Create a clean and tidy table with pre and post operation activity levels


## Load packages
We should only need the tidyverse for this exercise.

```{r load-packages}
library(tidyverse)
```

## Read in data

The data is in the file "Hip Replacement CCG 1819.csv", (which I have saved in the same folder as this rmd file) and it contains patient reported outcomes for hip replacement procedures, from April 2018 to March 2019. It was downloaded from https://digital.nhs.uk/data-and-information/publications/statistical/patient-reported-outcome-measures-proms/for-hip-and-knee-replacement-procedures-april-2018-to-march-2019
We also have the data dictionary for this dataset in "proms_data_dictionary.pdf".

```{r read-data}
hip_data <- read_csv("Hip Replacement CCG 1819.csv")
```

## Prepare the data

This includes three steps: 

inspecting the data, 
selecting only the variables we want, 
and dealing with missing values. 

```{r inspect-data}
glimpse(hip_data)
```

```{r select-gender-eq5d}
### Select gender and quality of life score pre and post operation

gender_EQ5D <- hip_data %>%
  select(Gender,`Pre-Op Q EQ5D Index`,`Post-Op Q EQ5D Index`) %>%
  rename(Sex  = `Gender`,
         EQ5D_Pre = `Pre-Op Q EQ5D Index`,
         EQ5D_Post = `Post-Op Q EQ5D Index`
         )

head(gender_EQ5D)
```

### Identify and remove missing values

```{r missing-values}
gender_EQ5D$Sex %>% unique()
gender_EQ5D$Sex %>% table()
gender_EQ5D %>% summary()

gender_EQ5D_noNA <- gender_EQ5D %>%
  drop_na() %>%
  filter(Sex !='*')

table(gender_EQ5D_noNA$Sex)
summary(gender_EQ5D_noNA)
```

### Replace 1 and 2 with "male" and "female"

```{r recode-gender}
gender_EQ5D_noNA <-gender_EQ5D_noNA %>%
  mutate(Sex = ifelse(Sex == 1, "male", "female"))

```


### Check that data is tidy
The data frame is not tidy, because the column names EQ5D_Pre and EQ5D_Post contain *data*: the time point when EQ5D was measured: pre or post operation.  

```{r tidy-data}
head(gender_EQ5D_noNA)

tidy_gender_EQ5D_noNA <- gender_EQ5D_noNA %>%
  pivot_longer(c(EQ5D_Pre,EQ5D_Post),
               names_to = 'Time',   # the name of the column to create from the data stored in the original column names
               names_prefix = 'EQ5D_',  # remove this text from the start of each variable name
               values_to = 'EQ5D' # the name of the column to create from the data stored in cell values
               )

head(tidy_gender_EQ5D_noNA)
  
```

## Plot quality of life pre and post operation for each sex group

```{r plot-eq5d}

# Turn Time into a "factor" so we can order the categories any way we want
# otherwise they are alphabetical and "Post" ends up before "Pre"
tidy_gender_EQ5D_noNA$Time <- factor(tidy_gender_EQ5D_noNA$Time,levels=c('Pre','Post'))

# ggplot creates a blank canvas, to which we add a boxplot with "geom_boxplot"
tidy_gender_EQ5D_noNA %>%
  ggplot() +
  geom_boxplot(aes(x = Time, y = EQ5D, colour = Sex))
```

## Question 2

2. Calculate how many patients in this dataset have been told by a doctor that they have problems caused by a stroke

### Select Stroke Column Only

```{r select-stroke-data}

stroke_data <- hip_data %>% 
  select(Stroke)


head(stroke_data)
```

### Calculate frequencies

```{r}
table(stroke_data)
```

### Interpreting Results

Using the data dictionary we can see that these values are interpreted 

Response to corresponding Q1 general health question:
Have you been told by a doctor that you have problems caused by a stroke?
Value
1 = Yes
9 = Missing

Therefore, we can interpret the results that 400 patients in this dataset
have been told they have problems caused by a stroke.

## Question 3

3. Create a clean and tidy table with pre and post operation activity levels

### Select activity pre and post operation

```{r select-activity-data}

activity_data <- hip_data %>% 
  select(`Pre-Op Q Activity`, `Post-Op Q Activity`) %>% 
  rename(Activity_Pre = `Pre-Op Q Activity`,
         Activity_Post = `Post-Op Q Activity`)

head(activity_data)
```

### remove missing values

```{r activity-missing-values}

activity_data %>% table()

activity_data_noNA <- activity_data %>% 
  filter(Activity_Pre != 9,
         Activity_Post != 9)

activity_data_noNA %>% table()

```

### Check data is tidy

```{r activity-tidy-data}
head(activity_data_noNA)

tidy_activity_data_noNA <- activity_data_noNA %>% 
  pivot_longer(c(Activity_Pre, Activity_Post),
               names_to = "Time",
               names_prefix = "Activity_",
               values_to = "Activity"
               )

head(tidy_activity_data_noNA)

```

